Foreground Signatures and Mitigation in Neutral Hydrogen Power Spectra From The Reionization Epoch

### 2014-10-23 foreground_delay_spectrum_modeling.tex
from d1904dc: ============
(*) Incorporated most of Josh's comments
(*) Changed one figure a little
(*) Changed the usage of "compact objects" -- addressing comments from both Josh
and Abraham

Detection of 21~cm emission of neutral hydrogen from the epoch of reionization, at redshifts $z>6$, is limited primarily by foreground emission.  We investigate the chromatic signatures of an all--sky foreground model imprinted by an instrumental transfer function.  Using the delay spectrum technique, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. Comparing data from recent Murchison Widefield Array observations with simulations separated into components based on type of emission, we identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact objects dominate the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier space.  Based on these results, we propose that selective down--weighting of baselines based on length, direction, and time will mitigate foreground contamination substantially (up to a factor $\sim 10$ with almost no loss of data) in redshifted 21~cm power spectrum analysis.

@@

Considerable effort has been and continues to be made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum and formulating robust estimators of the true power spectrum \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,thy13,liu14a,liu14b}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide--field (and chromatic) response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes discarded by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components at greatest risk to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

@@

The model described in the preceding section is primarily a model of the diffuse foreground sky. While it contains faint compact emission blended in with the diffuse emission, it is known to be devoid of bright compact sources \citep{deo08}. In order to supplement it with missing bright compact emission, we use classical radio source confusion estimates to determine the nominal flux density threshold and use compact sources above this threshold. It may be noted that slightly different criteria are in common use in radio astronomy to estimate radio source confusion \citep[see Appendix of][and references therein]{thy13}. For an angular resolution of 0\fdg 85, using a conservative `$S_\textrm{c}=5\sigma_\textrm{c}$' criterion, we determine the flux density threshold to be $\approx 10$~Jy. Other liberal criteria that yield a lower threshold carry a greater risk of double--counting compact sources which might be already blended in with the diffuse sky model.

We use a combination of the NRAO VLA Sky Survey \citep[NVSS;][]{con98} at 1.4~GHz and the Sydney University Molonglo Sky Survey \citep[SUMSS;][]{boc99,mau03} at 843~MHz to provide our compact object catalog due to their complementary survey footprints covering the entire sky, and matched flux density sensitivity and angular resolution. The SUMSS catalog covers the sky with declination $\delta < -30$\arcdeg~ with a limiting peak brightness of 6--10~mJy/beam and an angular resolution of $\sim 45$\arcsec. The NVSS covers the sky with $\delta > -40$\arcdeg~ with a similar angular resolution and a limiting flux density of $\approx 2.5$~mJy for compact objects.

From the SUMSS catalog, we select compact objects whose deconvolved major axes are equal to 0\arcsec. From the NVSS catalog, we excluded objects that overlap with those in the SUMSS survey footprint. Compact objects from NVSS were selected if the convolved major axes were not greater than $\approx 47$\arcsec, which is almost equal to the angular resolution of the survey. Using a mean spectral index of $\langle\alpha_\textrm{sp}\rangle=-0.83$ (flux density, $S(f)\propto f^{\alpha_\textrm{sp}}$) obtained by \citet{mau03} for both NVSS and SUMSS catalog objects, we calculate the corresponding flux densities at 185~MHz, $S_{185}$. From this subset, we choose compact objects with $S_{185}\geq 10$~Jy. The selection of such bright compact objects is not affected by minor differences in flux density sensitivity of the two surveys. We verified that our selection criteria ensure a similar areal density of objects in the two surveys.

Based on these criteria, we selected 133 objects from the SUMSS catalog and 336 objects from the NVSS catalog. Together with the diffuse foreground model, we obtain an all--sky foreground model consisting of both compact and diffuse emission. Figure~\ref{fig:CSM} shows the compact foreground emission model used in our study whose with contours of the power pattern overlaid. The flux densities of the compact sources are indicated by the color scale.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/csm.eps}
\caption{Flux densities of bright compact objects at 185~MHz visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels and corresponds to the color--coded filled circles. Contours of power--pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.001953125, 0.0078125, 0.03125, 0.125, and 0.5. The locations of bright compact objects appear to be random.\label{fig:CSM}}
\end{figure}

@@

In order to make a quantitative comparison of delay spectra obtained with MWA data and our simulations, we consider the uncertainty in the assumed spectral index of our foreground model. Our foreground models are derived from other higher frequency catalogs and sky maps. The inherent spread in spectral index increases the uncertainty while predicting fluxes at the observing frequency. Using simple error propagation, the fractional error in the delay spectrum caused by the spread in spectral index is $\sim \ln(f_\textrm{orig}/f)\,\Delta\alpha_\textrm{sp}$, where, $f_\textrm{orig}$ is the original frequency at which the catalog or map was created, $f=185$~MHz is the MWA observing frequency, and $\Delta\alpha_\textrm{sp}$ is the spread (HWHM) in spectral index. From \citet{mau03}, we assume $\Delta\alpha_\textrm{sp} \approx 0.35$ for compact objects from NVSS and SUMSS catalogs. Although the model of \citet{deo08} yields a spectral index per direction on the sky, we assume similar uncertainties exist in spectral indices of our diffuse sky model as well, which is predominantly derived from the 408~MHz map of \citet{has82}. Thus, fractional errors in delay spectrum from compact foreground objects and diffuse emission are $\sim$70\% and $\sim$30\% respectively.

@@

These uncertainties are presented to confirm the qualitative agreement already seen between Figures~\ref{fig:fhd_data} and \ref{fig:sim_data}. These uncertainty estimates are conservative. A full treatment of all uncertainties and deviations from ideal behavior such as frequency dependent errors in tile power--pattern \citep{ber14}, calibration \citep{dat10}, data corruption due to interference, anisoplanatic wide--field imaging and ionospheric effects \citep{int09} will bring the simulations much closer in agreement with observations. Our primary objective is to explore in detail the foreground signatures embedded in the {\it foreground wedge}.

@@

\item {\it Sky Model}: Our model of the sky consists of bright compact objects and diffuse emission on diverse spatial scales as shown in Figures~\ref{fig:DSM} and \ref{fig:CSM}. It also consists of localized regions of strong emission such as the Galactic plane and Galactic center. The resulting sky emission is anisotropic. In fact, the patch of sky for MWA observations is chosen from regions of low foreground emission. We will investigate the signatures of each component and their potential impact on {\it EoR window} contamination.

@@

Note that numerous features may overlap at different degrees of significance depending on combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as antenna spacing length and orientation, power--pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectra of foreground emission.

@@

Figure~\ref{fig:noiseless-dsm-delay-spectrum} shows the amplitude of delay spectrum, $|V_{b\tau}(\vec{b},\tau)|$, simulated with only the diffuse foreground model, omitting compact sources. The top and bottom panels correspond to {\it off--zenith} and {\it zenith} pointings, respectively.

@@

The amplitude response of an interferometer to a compact object is, to first order, flat with respect to baseline length. Since the main lobe of the power--pattern in the {\it off--zenith} pointing is centered eastward of zenith, the bulk of the compact foreground emission is seen in a branch with positive delay corresponding to the position of the primary lobe of the power--pattern. In the {\it zenith} pointing, compact emission from the same patch of sky is seen as a bright horizontal arm at zero delay  since the primary lobe of the power--pattern is centered at zenith.

Foreground emission in zero delay and negative delay regions in the {\it off--zenith} pointing is caused by compact objects co--located with secondary lobes of the power--pattern. On the other hand, compact objects co--located with secondary lobes of power--pattern in the {\it zenith} pointing are revealed as faint but distinct branches at positive and negative delays depending on the orientation of antenna spacing and direction of emission on the sky.

@@

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three antenna pairs of different antenna spacings oriented northward during the {\it zenith} pointing; each is a different vertical slice of the two dimensional delay spectrum plots shown in Figures~\ref{fig:fhd_data}, \ref{fig:sim_data}, \ref{fig:noiseless-dsm-delay-spectrum}, and \ref{fig:noiseless-csm-delay-spectrum}. The diffuse, compact, and composite components are shown as solid red, cyan, and black lines, respectively. The horizon delay limits are shown as a pair of vertical dotted lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. Uncertainty in emission in the {\it foreground} wedge (between the horizon delay limits) is dominated by the $\gtrsim$~70\% error owing to uncertainties in predicting the spectral index of compact foreground objects, while thermal noise fluctuations dominate outside.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_3_baseline_comparison_CLEAN_noiseless_visibilities_0.3m_ground_custom_gaussian_FG_model_all_sky_nside_64_Tsys_95.0K_185.0_MHz_30.7_MHz_bhw2.0.eps}
\caption{Simulated delay spectrum amplitudes $|V_{b\tau}(\vec{b},\tau)|$ for three chosen northward oriented antenna spacings of length: $\sim$84~m ({\it top}), $\sim$104~m ({\it middle}), and $\sim$171~m ({\it bottom}). The antenna spacing is specified in each panel along East, North and Upward directions. The solid red, cyan, and black lines denote contributions from diffuse emission, compact objects and composite foreground model respectively. Vertical dotted lines mark the horizon delay limits. The shaded region denotes the envelope of uncertainty around the composite model delay spectrum. This uncertainty is dictated by uncertainties in spectral indices of foreground models inside the delay horizon delay limits. Outside the horizon delay limits, it is predominantly due to thermal noise fluctuations. Compact objects dominate the central regions of delay spectrum while both components, especially diffuse emission on short antenna spacings, dominate near the horizon delay limits, giving rise to a {\it pitchfork}--shaped structure. \label{fig:pitchfork-baselines}}
\end{figure}

The zero--delay peak (corresponding to the main lobe in the power--pattern) with a value of $\sim 10^8$--$10^9$~Jy~Hz, independent of antenna spacing, is predominantly determined by compact foreground objects. The corresponding peak at zero--delay from diffuse foreground model is at least 10 times fainter and decreases drastically with an increase in antenna spacing. This is the response expected from different antenna spacings towards compact and diffuse foreground objects.

Near the horizon delay limits at shorter antenna spacing, diffuse emission is brighter than that from compact objects. Here, diffuse emission does not decrease as rapidly with increasing antenna spacing as was seen at zero--delay. In fact, even on widely spaced antennas, {\emph diffuse emission in the delay spectrum near the horizon delay limits exceeds that in the main lobe by more than two orders of magnitude}. We attribute this to apparent shortening of antenna spacings projected in the direction of diffuse emission entering far from the primary field of view, thereby giving rise to the edge--heavy {\it two--pronged fork} feature discussed in \S\ref{sec:diffuse-long-baselines}.

@@

The Galactic center in the {\it off--zenith} pointing is one such example already available in our study. Figure~\ref{fig:sky-model} shows the sky model ({\it top:} compact objects, {\it bottom:} diffuse sky model) in this pointing. The Galactic center is the most dominant source of foreground contamination from the diffuse sky model and is co--located with a bright secondary lobe of the power pattern near the western horizon. Figure~\ref{fig:delay-map} shows the sky mapped to delays registered by the baseline vectors, of length 100~m for instance, oriented northward (top panel) and eastward (bottom panel). Since the Galactic center is located in the western sky, this figure demonstrates that it is observed at negative delay on a baseline vector oriented eastward and at zero delay on a baseline vector oriented northward. Figure~\ref{fig:baseline-breakup} shows the delay spectra on baselines oriented northward (67\fdg 5~$\le \theta_\textrm{b} <$~112\fdg 5) at the top and eastward (-22\fdg 5~$\le \theta_\textrm{b} <$~22\fdg 5) at the bottom. The Galactic center manifests itself most distinctly near the negative horizon delay limit on short eastward baselines in the delay spectrum (bottom panel of Figure~\ref{fig:baseline-breakup}). Consequently, the spillover caused by the instrument's transfer function from the {\it foreground wedge} into the {\it EoR window} affects the northward baselines the least and is most severe in eastward baselines (particularly the short ones) evident by the bright vertical stripes of foreground contamination.

\begin{figure*}[htb]
\centering
\subfloat[][Sky model]{\label{fig:sky-model}\includegraphics[width=0.33\linewidth]{figures/v1_0/sky_model_snapshot_0.eps}}
\subfloat[][Delay maps]{\label{fig:delay-map}\includegraphics[width=0.33\linewidth]{figures/v1_0/directional_delay_map_snapshot_0.eps}}
\subfloat[][Delay Spectra]{\label{fig:baseline-breakup}\includegraphics[width=0.33\linewidth]{figures/v1_0/baseline_orientation_breakup_snapshot_0.eps}}
\caption{(a): Sky model showing compact objects ({\it top}) and diffuse emission ({\it bottom}). Contours of power pattern are overlaid whose levels are identical to those in Figures~\ref{fig:DSM} and \ref{fig:CSM}. The Galactic center is very prominent in diffuse emission in the western sky and is co--located with a bright secondary lobe of the power pattern. (b): Sky hemisphere mapped to delays observed on antenna spacings with northward ({\it top}) and eastward ({\it bottom}) orientations. Color scale shown is for a 100~m antenna spacing. Delays vary linearly with antenna spacing length. The bright Galactic center in the western sky is recorded at zero delays in northward antenna spacings and close to negative horizon delay limit on eastward antenna spacings. (c): Simulated delay spectra $|V_{b\tau}(\vec{b},\tau)|$ on antenna spacings oriented northward ({\it top}) and eastward ({\it bottom}). White lines denote horizon delay limits. The bright Galactic center is prominently visible close to negative horizon delay limit, especially on short eastward antenna spacings. These are also the most severely contaminated by foreground spillover. The northward antenna spacings, on the other hand, are the least contaminated.}
\label{fig:breakup}
\end{figure*}

@@

This screening technique can be generalized to optimize between foreground mitigation and loss of sensitivity from discarding data. Figure~\ref{fig:screening} shows how the typical foreground contamination\footnote{We use standard deviation of noiseless $V_{b\tau}(\vec{b},\tau)$ from foregrounds in the {\it EoR window} as a measure of foreground contamination.} in the {\it EoR window} depends on the orientations and lengths of discarded antenna spacings. We choose antenna spacings oriented eastward to varying degrees of directedness, i.e., -7\fdg 5~$\le\theta_\textrm{b}<$~7\fdg 5 (solid circles), -15\arcdeg~$\le\theta_\textrm{b}<$~15\arcdeg~ (solid squares), and -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5 (solid stars). Among antenna spacings that satisfy these criteria, we discard data from those whose lengths are shorter than $|\vec{b}|_\textrm{max}$ ($x$--axis) and show foreground contamination estimated in the {\it EoR window} from all remaining antenna spacings. In other words, these plots demonstrate the progress in mitigation as orientation and maximum length of discarded antenna spacings are varied. % Foreground contamination estimated from only northward (dashed line) and eastward (dot--dashed line) antenna spacings represent the lower and upper limits, respectively, and are shown for reference. 
The fraction of discarded antenna spacings discarded relative to the total number is shown in dotted lines for different ranges of $\theta_\textrm{b}$. It is seen that foreground contamination can be mitigated by a factor between $\sim 1.5$ ($|\theta_\textrm{b}|\le$~7\fdg 5) and $\sim 10$ ($|\theta_\textrm{b}|\le$~22\fdg 5). It is notable that short antenna spacings play maximum role in determining foreground contamination. For instance, among antenna spacings with orientations $|\theta_\textrm{b}|\le$~22\fdg 5, discarding those with lengths $|\vec{b}|\gtrsim 30$~m does not mitigate foreground contamination any further and would only lead to loss of sensitivity as the fraction of discarded baselines increases from $\sim 5$\% to $\sim 25$\%. On the other hand, a bright compact object at the same location will give rise to foreground contamination even on longer antenna spacings. Such cases will necessitate discarding more or, in the worst case, all of the eastward oriented antenna spacings accounting for $\sim 25$\% loss of data.

@@

Simulating in all important respects the response of the MWA to an all--sky foreground model that consists of diffuse Galactic emission from \citet{deo08} and bright compact objects from the NVSS and SUMSS catalogs, we confirm that the modeled delay spectra are in agreement with data obtained with the MWA, to the extent allowed by levels of uncertainty known in the foreground models and thermal noise fluctuations in measurements.

>>>>

Detection of 21~cm emission of neutral hydrogen from the epoch of reionization, at redshifts $z>6$, is limited primarily by foreground emission.  We investigate the chromatic signatures of an all--sky foreground model imprinted by an instrumental transfer function.  Using the delay spectrum technique, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. Comparing data from recent Murchison Widefield Array observations with simulations separated into components based on type of emission, we identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact emission dominates the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier space. Based on these results, we propose that selective down--weighting of baselines based on length, direction, and time will mitigate foreground contamination substantially (up to a factor $\sim 10$ with almost no loss of data) in redshifted 21~cm power spectrum analysis.

@@

Considerable effort has been and continues to be made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum and formulating robust estimators of the true power spectrum \citep{bow09,liu09,dat10,liu11,mor12,tro12,pob13,thy13,dil14,liu14a,liu14b}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide--field (and chromatic) response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes discarded by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components at greatest risk to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

@@

The model described in the preceding section is primarily a model of the diffuse foreground sky. While it contains faint compact emission blended in with the diffuse emission, it is known to be devoid of bright point sources \citep{deo08}. In order to supplement it with missing bright compact emission, we use classical radio source confusion estimates to determine the nominal flux density threshold and include point sources brighter than this threshold. It may be noted that slightly different criteria are in common use in radio astronomy to estimate radio source confusion \citep[see Appendix of][and references therein]{thy13}. For an angular resolution of 0\fdg 85, using a conservative `$S_\textrm{c}=5\sigma_\textrm{c}$' criterion, we determine the flux density threshold to be $\approx 10$~Jy. Other liberal criteria that yield a lower threshold carry a greater risk of double--counting point sources which might be already blended in with the diffuse sky model.

We use a combination of the NRAO VLA Sky Survey \citep[NVSS;][]{con98} at 1.4~GHz and the Sydney University Molonglo Sky Survey \citep[SUMSS;][]{boc99,mau03} at 843~MHz to provide our point source catalog due to their complementary survey footprints covering the entire sky, and matched flux density sensitivity and angular resolution. The SUMSS catalog covers the sky with declination $\delta < -30$\arcdeg~ with a limiting peak brightness of 6--10~mJy/beam and an angular resolution of $\sim 45$\arcsec. The NVSS covers the sky with $\delta > -40$\arcdeg~ with a similar angular resolution and a limiting flux density of $\approx 2.5$~mJy for point sources.

From the SUMSS catalog, we select point sources whose deconvolved major axes are equal to 0\arcsec. From the NVSS catalog, we excluded objects that overlap with those in the SUMSS survey footprint. Point sources from NVSS were selected if the convolved major axes were not greater than $\approx 47$\arcsec, which is almost equal to the angular resolution of the survey. Using a mean spectral index of $\langle\alpha_\textrm{sp}\rangle=-0.83$ (flux density, $S(f)\propto f^{\alpha_\textrm{sp}}$) obtained by \citet{mau03} for both NVSS and SUMSS catalog objects, we calculate the corresponding flux densities at 185~MHz, $S_{185}$. From this subset, we choose point sources with $S_{185}\geq 10$~Jy. The selection of such bright point sources is not affected by minor differences in flux density sensitivity of the two surveys. We verified that our selection criteria ensure a similar areal density of objects in the two surveys.

Based on these criteria, we selected 133 objects from the SUMSS catalog and 336 objects from the NVSS catalog. Together with the diffuse foreground model, we obtain an all--sky foreground model consisting of both compact and diffuse emission. Figure~\ref{fig:CSM} shows the compact foreground emission model used in our study whose with contours of the power pattern overlaid. The flux densities of the point sources are indicated by the color scale.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/csm.eps}
\caption{Flux densities of bright point sources at 185~MHz visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels and corresponds to the color--coded filled circles. Contours of power--pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.001953125, 0.0078125, 0.03125, 0.125, and 0.5. \label{fig:CSM}}
\end{figure}

@@

In order to make a quantitative comparison of delay spectra obtained with MWA data and our simulations, we consider the uncertainty in the assumed spectral index of our foreground model. Our foreground models are derived from other higher frequency catalogs and sky maps. The inherent spread in spectral index increases the uncertainty while predicting fluxes at the observing frequency. Using simple error propagation, the fractional error in the delay spectrum caused by the spread in spectral index is $\sim \ln(f_\textrm{orig}/f)\,\Delta\alpha_\textrm{sp}$, where, $f_\textrm{orig}$ is the original frequency at which the catalog or map was created, $f=185$~MHz is the MWA observing frequency, and $\Delta\alpha_\textrm{sp}$ is the spread (HWHM) in spectral index. From \citet{mau03}, we assume $\Delta\alpha_\textrm{sp} \approx 0.35$ for point sources from NVSS and SUMSS catalogs. Although the model of \citet{deo08} yields a spectral index per direction on the sky, we could assume similar uncertainties exist in spectral indices of our diffuse sky model as well, which is predominantly derived from the 408~MHz map of \citet{has82}. Thus, fractional errors in delay spectrum from compact and diffuse components are $\sim$70\% and $\sim$30\% respectively.

@@

These uncertainties are presented to confirm the qualitative agreement already seen between Figures~\ref{fig:fhd_data} and \ref{fig:sim_data}. These uncertainty estimates are conservative. A full treatment of all uncertainties and deviations from ideal behavior such as frequency dependent errors in tile power--pattern \citep{ber14}, calibration \citep{dat10}, data corruption due to interference, anisoplanatic wide--field imaging and ionospheric effects \citep{int09} will bring the simulations much closer in agreement with observations, but is beyond the scope of this paper. Our primary objective is to explore in detail the foreground signatures embedded in the {\it foreground wedge}.

@@

\item {\it Sky Model}: Our model of the sky consists of compact and diffuse emission on diverse spatial scales as shown in Figures~\ref{fig:DSM} and \ref{fig:CSM}. It also consists of localized regions of strong emission such as the Galactic plane and Galactic center. The resulting sky emission is anisotropic. In fact, the patch of sky for MWA observations is chosen from regions of low foreground emission. We will investigate the signatures of each component and their potential impact on {\it EoR window} contamination.

@@

In subsequent sections, we provide a detailed explanation of our results as a combination of factors noted above. Note that numerous features may overlap at different degrees of significance depending on combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as antenna spacing length and orientation, power--pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectra of foreground emission.

@@

Figure~\ref{fig:noiseless-dsm-delay-spectrum} shows the amplitude of delay spectrum, $|V_{b\tau}(\vec{b},\tau)|$, simulated with only the diffuse foreground model. The top and bottom panels correspond to {\it off--zenith} and {\it zenith} pointings, respectively.

@@

The amplitude response of an interferometer to a point source is, to first order, flat with respect to baseline length. Since the main lobe of the power--pattern in the {\it off--zenith} pointing is centered eastward of zenith, the bulk of the compact foreground emission is seen in a branch with positive delay corresponding to the position of the primary lobe of the power--pattern. In the {\it zenith} pointing, compact emission from the same patch of sky is seen as a bright horizontal arm at zero delay  since the primary lobe of the power--pattern is centered at zenith.

Foreground emission in zero delay and negative delay regions in the {\it off--zenith} pointing is caused by point sources co--located with secondary lobes of the power--pattern. On the other hand, point sources co--located with secondary lobes of power--pattern in the {\it zenith} pointing are revealed as faint but distinct branches at positive and negative delays depending on the orientation of antenna spacing and direction of emission on the sky.

@@

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three antenna pairs of different antenna spacings oriented northward during the {\it zenith} pointing; each is a different vertical slice of the two dimensional delay spectrum plots shown in Figures~\ref{fig:fhd_data}, \ref{fig:sim_data}, \ref{fig:noiseless-dsm-delay-spectrum}, and \ref{fig:noiseless-csm-delay-spectrum}. The diffuse, compact, and composite components are shown as solid red, cyan, and black lines, respectively. The horizon delay limits are shown as a pair of vertical dotted lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. Uncertainty in emission in the {\it foreground} wedge (between the horizon delay limits) is dominated by the $\gtrsim$~70\% error owing to uncertainties in predicting the spectral index of compact foreground model, while thermal noise fluctuations dominate outside.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_3_baseline_comparison_CLEAN_noiseless_visibilities_0.3m_ground_custom_gaussian_FG_model_all_sky_nside_64_Tsys_95.0K_185.0_MHz_30.7_MHz_bhw2.0.eps}
\caption{Simulated delay spectrum amplitudes $|V_{b\tau}(\vec{b},\tau)|$ for three chosen northward oriented antenna spacings of length: $\sim$84~m ({\it top}), $\sim$104~m ({\it middle}), and $\sim$171~m ({\it bottom}). The antenna spacing is specified in each panel along East, North and Upward directions. The solid red, cyan, and black lines denote contributions from diffuse, compact, and composite foreground models respectively. Vertical dotted lines mark the horizon delay limits. The shaded region denotes the envelope of uncertainty around the composite model delay spectrum. This uncertainty is dictated by uncertainties in spectral indices of foreground models inside the delay horizon delay limits. Outside the horizon delay limits, it is predominantly due to thermal noise fluctuations. Compact emission dominates the central regions of delay spectrum while both components, especially diffuse emission on short antenna spacings, dominate near the horizon delay limits, giving rise to a {\it pitchfork}--shaped structure. \label{fig:pitchfork-baselines}}
\end{figure}

The zero--delay peak (corresponding to the main lobe in the power--pattern) with a value of $\sim 10^8$--$10^9$~Jy~Hz, independent of antenna spacing, is predominantly determined by compact emission. The corresponding peak at zero--delay from diffuse foreground model is at least 10 times fainter and decreases drastically with an increase in antenna spacing. This is the response expected from different antenna spacings towards compact and diffuse emission.

Near the horizon delay limits at shorter antenna spacing, the diffuse component is brighter relative to the compact component. Here, diffuse emission does not decrease as rapidly with increasing antenna spacing as was seen at zero--delay. In fact, even on widely spaced antennas, {\emph diffuse emission in the delay spectrum near the horizon delay limits exceeds that in the main lobe by more than two orders of magnitude}. We attribute this to apparent shortening of antenna spacings projected in the direction of diffuse emission entering far from the primary field of view, thereby giving rise to the edge--heavy {\it two--pronged fork} feature discussed in \S\ref{sec:diffuse-long-baselines}.

@@

The Galactic center in the {\it off--zenith} pointing is one such example already available in our study. Figure~\ref{fig:sky-model} shows the sky model ({\it top:} compact component, {\it bottom:} diffuse component) in this pointing. The Galactic center is the most dominant source of foreground contamination from the diffuse sky model and is co--located with a bright secondary lobe of the power pattern near the western horizon. Figure~\ref{fig:delay-map} shows the sky mapped to delays registered by the baseline vectors, of length 100~m for instance, oriented northward (top panel) and eastward (bottom panel). Since the Galactic center is located in the western sky, this figure demonstrates that it is observed at negative delay on a baseline vector oriented eastward and at zero delay on a baseline vector oriented northward. Figure~\ref{fig:baseline-breakup} shows the delay spectra on baselines oriented northward (67\fdg 5~$\le \theta_\textrm{b} <$~112\fdg 5) at the top and eastward (-22\fdg 5~$\le \theta_\textrm{b} <$~22\fdg 5) at the bottom. The Galactic center manifests itself most distinctly near the negative horizon delay limit on short eastward baselines in the delay spectrum (bottom panel of Figure~\ref{fig:baseline-breakup}). Consequently, the spillover caused by the instrument's transfer function from the {\it foreground wedge} into the {\it EoR window} affects the northward baselines the least and is most severe in eastward baselines (particularly the short ones) evident by the bright vertical stripes of foreground contamination.

\begin{figure*}[htb]
\centering
\subfloat[][Sky model]{\label{fig:sky-model}\includegraphics[width=0.33\linewidth]{figures/v1_0/sky_model_snapshot_0.eps}}
\subfloat[][Delay maps]{\label{fig:delay-map}\includegraphics[width=0.33\linewidth]{figures/v1_0/directional_delay_map_snapshot_0.eps}}
\subfloat[][Delay Spectra]{\label{fig:baseline-breakup}\includegraphics[width=0.33\linewidth]{figures/v1_0/baseline_orientation_breakup_snapshot_0.eps}}
\caption{(a): Sky model showing compact ({\it top}) and diffuse ({\it bottom}) emission. Contours of power pattern are overlaid whose levels are identical to those in Figures~\ref{fig:DSM} and \ref{fig:CSM}. The Galactic center is very prominent in diffuse emission in the western sky and is co--located with a bright secondary lobe of the power pattern. (b): Sky hemisphere mapped to delays observed on antenna spacings with northward ({\it top}) and eastward ({\it bottom}) orientations. Color scale shown is for a 100~m antenna spacing. Delays vary linearly with antenna spacing length. The bright Galactic center in the western sky is recorded at zero delays in northward antenna spacings and close to negative horizon delay limit on eastward antenna spacings. (c): Simulated delay spectra $|V_{b\tau}(\vec{b},\tau)|$ on antenna spacings oriented northward ({\it top}) and eastward ({\it bottom}). White lines denote horizon delay limits. The bright Galactic center is prominently visible close to negative horizon delay limit, especially on short eastward antenna spacings. These are also the most severely contaminated by foreground spillover. The northward antenna spacings, on the other hand, are the least contaminated.}
\label{fig:breakup}
\end{figure*}

@@

This screening technique can be generalized to optimize between foreground mitigation and loss of sensitivity from discarding data. Figure~\ref{fig:screening} shows how the typical foreground contamination\footnote{We use standard deviation of noiseless $V_{b\tau}(\vec{b},\tau)$ from foregrounds in the {\it EoR window} as a measure of foreground contamination.} in the {\it EoR window} depends on the orientations and lengths of discarded antenna spacings. We choose antenna spacings oriented eastward to varying degrees of directedness, i.e., -7\fdg 5~$\le\theta_\textrm{b}<$~7\fdg 5 (solid circles), -15\arcdeg~$\le\theta_\textrm{b}<$~15\arcdeg~ (solid squares), and -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5 (solid stars). Among antenna spacings that satisfy these criteria, we discard data from those whose lengths are shorter than $|\vec{b}|_\textrm{max}$ ($x$--axis) and show foreground contamination estimated in the {\it EoR window} from all remaining antenna spacings. In other words, these plots demonstrate the progress in mitigation as orientation and maximum length of discarded antenna spacings are varied. % Foreground contamination estimated from only northward (dashed line) and eastward (dot--dashed line) antenna spacings represent the lower and upper limits, respectively, and are shown for reference. 
The fraction of discarded antenna spacings discarded relative to the total number is shown in dotted lines for different ranges of $\theta_\textrm{b}$. It is seen that foreground contamination can be mitigated by a factor between $\sim 1.5$ ($|\theta_\textrm{b}|\le$~7\fdg 5) and $\sim 10$ ($|\theta_\textrm{b}|\le$~22\fdg 5). It is notable that short antenna spacings play maximum role in determining foreground contamination. For instance, among antenna spacings with orientations $|\theta_\textrm{b}|\le$~22\fdg 5, discarding those with lengths $|\vec{b}|\gtrsim 30$~m does not mitigate foreground contamination any further and would only lead to loss of sensitivity as the fraction of discarded baselines increases from $\sim 5$\% to $\sim 25$\%. On the other hand, a bright point source at the same location will give rise to foreground contamination even on longer antenna spacings. Such cases will necessitate discarding more or, in the worst case, all of the eastward oriented antenna spacings accounting for $\sim 25$\% loss of data.

@@

Simulating in all important respects the response of the MWA to an all--sky foreground model that consists of diffuse Galactic emission from \citet{deo08} and bright point sources from the NVSS and SUMSS catalogs, we confirm that the modeled delay spectra are in agreement with data obtained with the MWA, to the extent allowed by levels of uncertainty known in the foreground models and thermal noise fluctuations in measurements.
